---
title: "Analysis of Guinea Pig Tooth Growth"
author: "Steve Myles"
date: "July 22, 2015"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 2
---

# Data Summary / Exploratory Analysis

This report provides a comparison of guinea pig odontoblast (tooth) growth 
(`len`, measured in microns) by supplement (`supp`) and daily dosage (`dose`) 
using the `ToothGrowth` data from the R `datasets` package.  First, the data 
were loaded into a data frame (replacing `dose`'s values with factors as no
calculations were performed on dosage) and relevant packages were loaded as shown
in Appendix 1. The data were then evaluated and summarized.

```{r data summary, echo=FALSE}
## effectively turn off scientific notation
options(scipen=999)
## load packages and data, convert dose to factor
library(dplyr, warn.conflicts = FALSE) ## load the dplyr package
library(lattice) ## load the lattice package
data <- ToothGrowth ## assign the ToothGrowth data to a data frame
data$dose <- as.factor(data$dose) ## convert dose to factor
```

This data set has 60 observations of 3 variables (`len`, tooth length in 
microns; `supp`, either "OJ" for orange juice or "VC" for an ascorbic acid 
vitamin C supplement; and `dose`, the daily dosage of vitamin C, either 0.5mg, 
1mg, or 2mg).  Each combination of `supp` and `dose` has 10 observations so 
there are 30 total observations for each `supp` and 20 for each `dose`.  A data 
summary using the `str` and `summary` functions can be found in Appendix 2.

The data were plotted in a series of panel histograms (one for each `dose` 
regardless of `supp`, one for each `supp` regardless of `dose`, and one for each
combination of `supp` and `dose`) using the `lattice` package's `histogram` 
function.  Code for producing these plots is available in Appendix 3.

```{r exploratory plots, fig.width=3, fig.height=2, echo=FALSE}
## histogram of length by dose
histogram(~len | dose, data = data, layout = c(3, 1), type = "count",
          main = "Fig. 1: len by dose", xlab = "length")
```
\newline
```{r exploratory plots2, fig.width=3, fig.height=3, echo=FALSE}
## histogram of length by supp
histogram(~len | supp, data = data, layout = c(1, 2), type = "count",
          main = "Fig. 2: len by supp", xlab = "length")
## histogram of length by combination of dose and supplement
histogram(~len | dose * supp, data = data, layout = c(3, 2), type = "count",
          main = "Fig. 3: len by supp and dose", xlab = "length")
```

From the plots, it appears that higher dosage yields larger lengths (Fig. 1), 
regardless of supplement (in Fig. 3, `len` seems to be greater for `dose` = 2 
for both `supp` = "VC" and "OJ").  Additionally, in Fig. 3, orange juice ("OJ") 
seems to be associated with larger lengths than ascorbic acid ("VC"), regardless 
of dosage (the distribution of length for "OJ" (Fig. 2, as well as for each 
dosage in Fig. 3) seems to be further to the right than the distribtuion of 
length for "VC."  However, the relationship of `supp` and `len` is less clear in
Fig. 2 where the two distributions overlap to a large extent.

The means and standard deviations for each level of `supp` and `dose` and each 
combination of these factors were calculated using the `dplyr` package's 
`summarize` function.  Code for generating this data frame is available in 
Appendix 4.

```{r summary statistics, echo = FALSE}
## summarize by dose
summDose <- summarize(group_by(data, dose), mean = mean(len), sd = sd(len), 
                      n = n())
summDose$supp <- "total"
## summarize by supplement
summSupp <- summarize(group_by(data, supp), mean = mean(len), sd = sd(len), 
                      n = n())
summSupp$dose <- "total"
## summarize by dose and supplement
summDoseSupp <- summarize(group_by(data, dose, supp), mean = mean(len), 
                          sd = sd(len), n = n())

## combine summaries and sort using dplyr's arrange function
summ <- data.frame()
summ <- rbind(summ, summDose, summSupp, summDoseSupp)
summ <- arrange(summ, dose, supp)

## display the summary data frame
message("summary data frame:  'total' refers to either dose irrespective of
        supplement or supplement irrespective of dose")
summ
```

As can be seen, the sample means differ.  Tests were then performed to determine
conclusions about the population means based on these samples.

\newpage

# Hypothesis Tests

The `data` data frame was split into 11 data frames (one corresponding to each 
row of the above summary) for easy t-testing of each combination of `dose` and 
`supp`.  Seven t-tests with $\alpha$ = 0.05 were performed to compare the mean 
lengths of each group (i.e., comparing the mean `len` for the values of `dose` 
and `supp` as well as for each of their combinations -- a `dose` of 0.5mg was 
compared to a `dose` of 1mg and a `dose` of 2mg irrespective of `supp`, "VC" was 
compared to "OJ" irrespective of dose, and "VC" and "OJ" were compared for each 
level of `dose`).  Due to the small sample sizes, t-tests were chosen for 
comparing the distribution means.  Null hypotheses, alternative hypotheses, and 
$p$-values are provided below for each hypothesis test and the code used to 
subset and t-tests the groups is available in Appendix 5.

```{r subsets and t-tests, echo=FALSE}
## ***** subsetting *****
## 
## subset the data frame based on dose, supplement, and combinations of them
## rows where `dose` = "0.5" (irrespective of `supp`)
data0.5 <- data[data$dose=="0.5",]
## rows where `dose` = "1" (irrespective of `supp`)
data1 <- data[data$dose=="1",]
## rows where `dose` = "2" (irrespective of `supp`)
data2 <- data[data$dose=="2",]
## rows where `supp` = "VC" (irrespective of `dose`)
dataVC <- data[data$supp=="VC",]
## rows where `supp` = "OJ" (irrespective of `dose`)
dataOJ <- data[data$supp=="OJ",]
## rows where `dose` = "0.5" and `supp` = "VC"
data0.5VC <- data[data$dose == "0.5" & data$supp=="VC",]
## rows where `dose` = "0.5" and `supp` = "OJ"
data0.5OJ <- data[data$dose == "0.5" & data$supp=="OJ",]
## rows where `dose` = "1" and `supp` = "VC"
data1VC <- data[data$dose == "1" & data$supp=="VC",]
## rows where `dose` = "1" and `supp` = "OJ"
data1OJ <- data[data$dose == "1" & data$supp=="OJ",]
## rows where `dose` = "2" and `supp` = "VC"
data2VC <- data[data$dose == "2" & data$supp=="VC",]
## rows where `dose` = "2" and `supp` = "OJ"
data2OJ <- data[data$dose == "2" & data$supp=="OJ",]

## ***** t-tests with p-values displayed under hypothesis tests ******
##
## t-test for means of dose = 0.5 and dose = 1
tdose0.5_1 <- t.test(data0.5$len, data1$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pval0.5_1 <- ifelse(tdose0.5_1$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tdose0.5_1$p.value, digits = 5)))

## t-test for means of dose = 0.5 and dose = 2
tdose0.5_2 <- t.test(data0.5$len, data2$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pval0.5_2 <- ifelse(tdose0.5_2$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tdose0.5_2$p.value, digits=5)))

## t-test for means of dose = 1 and dose = 2
tdose1_2 <- t.test(data1$len, data2$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pval1_2 <- ifelse(tdose1_2$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tdose1_2$p.value, digits = 5)))

## t-test for means of supp = VC and supp = OJ
tsupp <- t.test(dataVC$len, dataOJ$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pvalsupp <- ifelse(tsupp$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tsupp$p.value, digits = 5)))

## t-test for means of dose = 0.5 for supp = VC and supp = OJ
tdose0.5 <- t.test(data0.5VC$len, data0.5OJ$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pval0.5 <- ifelse(tdose0.5$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tdose0.5$p.value, digits = 5)))

## t-test for means of dose = 1 for supp = VC and supp = OJ
tdose1 <- t.test(data1VC$len, data1OJ$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pval1 <- ifelse(tdose1$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tdose1$p.value, digits = 5)))

## t-test for means of dose = 2 for supp = VC and supp = OJ
tdose2 <- t.test(data2VC$len, data2OJ$len)
## determine p-value and display "< 0.00001" if it is or "= p-value" if not
pval2 <- ifelse(tdose2$p.value < 0.00001, paste0("< 0.00001"), 
                    paste0("= ", round(tdose2$p.value, digits = 5)))
```
### Hypothesis tests and associated p-values:

1. $H_0: \mu_{dose=0.5} = \mu_{dose=1}$  
$H_A: \mu_{dose=0.5} \neq \mu_{dose=1}$  
$p$-value `r pval0.5_1`
2. $H_0: \mu_{dose=0.5} = \mu_{dose=2}$  
$H_A: \mu_{dose=0.5} \neq \mu_{dose=2}$  
$p$-value `r pval0.5_2`
3. $H_0: \mu_{dose=1} = \mu_{dose=2}$  
$H_A: \mu_{dose=1} \neq \mu_{dose=2}$  
$p$-value `r pval1_2`
4. $H_0: \mu_{supp=VC} = \mu_{supp=OJ}$  
$H_A: \mu_{supp=VC} \neq \mu_{supp=OJ}$  
$p$-value `r pvalsupp`
5. $H_0: \mu_{dose=0.5,supp=VC} = \mu_{dose=0.5,supp=OJ}$  
$H_A: \mu_{dose=0.5,supp=VC} \neq \mu_{dose=0.5,supp=OJ}$  
$p$-value `r pval0.5`
6. $H_0: \mu_{dose=1,supp=VC} = \mu_{dose=1,supp=OJ}$  
$H_A: \mu_{dose=1,supp=VC} \neq \mu_{dose=1,supp=OJ}$  
$p$-value `r pval1`
7. $H_0: \mu_{dose=2,supp=VC} = \mu_{dose=2,supp=OJ}$  
$H_A: \mu_{dose=2,supp=VC} \neq \mu_{dose=2,supp=OJ}$  
$p$-value `r pval2`

# Conclusions

Based on the sample data, the null hypotheses ($H_0$) are accepted for cases 4 
(t-test comparing mean `len` of supplements "VC" and "OJ" regardless of `dose`) 
and 7 (t-test comparing mean `len` of supplements "VC" and "OJ" for `dose` = 
2mg) as these t-tests' $p$-values > $\alpha$ = 0.05.  $H_0$ in cases 1-3 and 5-6 
must be rejected in favor of the alternative hypotheses ($H_A$) because their 
$p$-values < $\alpha$ = 0.05.  

In other words, when `supp` is not a factor, the individual values of `dose` are 
associated with different average odontoblast lengths in microns (as evidenced 
by hypothesis tests 1-3).  Additionally, based on hypothesis test 4, one cannot 
say that the different supplements (`supp`, orange juice ("OJ") and ascorbic 
acid ("VC")) are associated with different mean odontoblast lengths.  However, 
taking both `dose` and `supp` into account yields different results:  when 
`dose` = 0.5mg/day and when `dose` = 1mg/day, one cannot say that the mean 
odontoblast lengths are different for the different supplements (hypothesis 
tests 5-6) but when `dose` = 2mg/day, one can assume that the mean length is the
same regardless of supplement (as $H_0$ is accepted in that case).

\newpage

# Appendix

## 1: Loading the data and required packages

```{r data summary, echo=TRUE, eval=FALSE}
```

## 2: Structure of the `ToothGrowth` data after assigning to the `data` data frame

```{r appendix}
str(data) ## review the structure of the data frame
summary(data)
```

## 3: Code for exploratory data analysis

```{r exploratory plots, echo=TRUE, eval=FALSE}
```

```{r exploratory plots2, echo=TRUE, eval=FALSE}
```
\newpage

## 4: Code for summary statistics

```{r summary statistics, echo=TRUE, eval=FALSE}
```

## 5: Code for subsetting and t-testing the data

```{r subsets and t-tests, echo=TRUE, eval=FALSE}
```